---
title: "Module 6 Final Project - Final Report"
subtitle: "ALY6040: Data Mining"
author: "Cheng Liu"
date: "Oct.25.2024"
output:
  pdf_document:
    fig_caption: yes
    toc: true
    number_sections: true
    highlight: tango
header-includes:
  - \usepackage{xcolor}
fontsize: 12pt
linestretch: 1.5 
mainfont: "Times New Roman"
geometry: margin=1in
---

# Introduction  
  The dataset for this final project contains detailed information on gun violence in the United
States from January 2013 to March 2018. The data can help identify trends and patterns in gun violence. By analyzing the data, I hope to use data mining techniques to find important insights that may also help provide effective recommendations to promote a safer society.  

  *Here are my business questions from Module 3:*  
  **1.** What factors predict severe incidents?  
  **2.** Are there times of the year when incidents are more likely?  
  **3.** Which incidents are most likely to have multiple casualties?  
  **4.** How do local factors relate to high incident rates in certain states?  
  
  To answer these questions, I used a dataset containing details about gun violence incidents, including participant roles, locations, and characteristics. This analysis involves cleaning the data, exploring trends, and building models to predict high-casualty incidents.

# Analysis

## Data Preparation  

### Data Cleaning:  
  Removed missing and unneeded columns to make the data consistent.  
  
### Feature Engineering:  
  I combined the number of kills and injuries into a new feature called `total casualties` to represent the severity of each incident, and created a binary variable `high_casualty` to identify incidents with three or more casualties. I also use one-hot encoding to convert some non-numeric columns into `0` or `1`.  
    
### Encoding:  
  Converted categorical features like `state`, `month`, and `weekday` into numbers to work with machine learning models.

## Methodologies  

### Tools:  
  I used R and several libraries like dplyr, ggplot2, and randomForest.  
  
### Techniques:  
  **Exploratory Data Analysis (EDA):** I looked at the most common ages, genders, and time trand involved in gun violence as well as the word cloud with the most common location.  
  **Machine Learning Model:** I used a `Random Forest` classifier to predict high-casualty incidents. This model helps to understand which features are most important in determining severe cases. (I created Logistic Regression, LightGBM, and K-Nearest Neighbors models, but due to the long processing times and unsatisfactory results, I removed the code and kept only a summary of the final results in Model Comparison part.)  
  
# R Code for the Whole Gun Violence Analysis  

## Code for EDA
```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(lubridate)
library(gridExtra)
library(ggrepel)
library(maps)
library(scales)
library(splitstackshape)

# Loading the dataset as a dataframe
gun_data <- read.csv("C:/data/raw/gun-violence-data_01-2013_03-2018.csv")

# Summary of the dataset
summary(gun_data)

# Checking data types of each column
data_types <- sapply(gun_data, class)
data_types_df <- tibble(Column = names(data_types), DataType = data_types)

# Counting unique values in each column
unique_counts <- sapply(gun_data, function(x) length(unique(x)))
unique_counts_df <- tibble(Column = names(unique_counts), UniqueValues = unique_counts)

# Combining data types and unique counts
data_summary <- left_join(data_types_df, unique_counts_df, by = "Column") %>%
  arrange(UniqueValues)
data_summary

# Finding and handling missing data
missing_data <- sapply(gun_data, function(x) sum(is.na(x)))
missing_percentage <- (missing_data / nrow(gun_data)) * 100
missing_summary <- tibble(
  Column = names(missing_data),
  MissingCount = missing_data,
  MissingPercentage = missing_percentage
) %>%
  arrange(desc(MissingPercentage))
missing_summary

# Detecting and removing duplicates
duplicate_count <- sum(duplicated(gun_data))
cat("Number of duplicate rows:", duplicate_count, "\n")

# Visualizing outliers in 'n_killed' and 'n_injured'
ggplot(gun_data, aes(y = n_killed)) +
  geom_boxplot(fill = "blue", color = "black", alpha = 0.7) +
  labs(title = "Outliers in n_killed", y = "Number of People Killed") +
  theme_minimal()

ggplot(gun_data, aes(y = n_injured)) +
  geom_boxplot(fill = "red", color = "black", alpha = 0.7) +
  labs(title = "Outliers in n_injured", y = "Number of People Injured") +
  theme_minimal()

# Identifying relationships between 'n_killed' and 'n_injured'
ggplot(gun_data, aes(x = n_killed, y = n_injured)) +
  geom_point(alpha = 0.5, color = "darkgreen") +
  labs(title = "Relationship between Number Killed and Number Injured", x = "Number of People Killed", y = "Number of People Injured") +
  theme_minimal()
```  

### Data Preparation Summary  

  **1.** Data Types and Unique Values: Each column’s data type and unique values were checked to aid analysis.  
  
  **2.** Handling Missing Data: High missing values in columns like `n_guns_involved` may impact modeling.  
  
  **3.** Duplicate Data: No duplicate rows found, indicating clean data. 
  
  **Summary:** These steps ensure data quality and provide a foundation for further analysis and modeling.  
  
  **4.** `n_killed` and `n_injured`: Most incidents have low casualties, but a few show extreme high numbers of deaths and injuries.  
  
  **5.** Relationship Between `n_killed` and `n_injured` Correlation: There’s a positive correlation between deaths and injuries, suggesting combining them as `total casualties` for a better measure of severity.

```{r, message=FALSE, warning=FALSE}
# Bar chart for top 10 states by incident frequency
top_states <- gun_data %>%
  count(state, sort = TRUE) %>%
  slice_max(n, n = 10)

ggplot(top_states, aes(x = reorder(state, n), y = n)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(title = "Top 10 States by Incident Frequency", x = "State", y = "Number of Incidents") +
  theme_minimal()
```

### Insight:  

  Illinois, California, and Florida have the highest gun violence incident frequencies, indicating these states could benefit most from focused interventions.

```{r, message=FALSE, warning=FALSE}
# Extracting year and quarter
gun_data$date <- ymd(gun_data$date)
gun_data <- gun_data %>% mutate(year = year(date), quarter = quarter(date))

# Plotting number of incidents by year
gun_data %>%
  ggplot(aes(x = as.factor(year))) +
  geom_bar(stat = 'count', fill = 'red') +
  scale_y_continuous(labels = comma) +
  geom_label(stat = "count", aes(label = after_stat(count), y = after_stat(count))) +
  labs(x = 'Year', y = 'Number of Incidents', title = 'Number of Incidents by Year (2013-2018)')
```

### Insight:  

  Gun violence incidents increased yearly from `2014` to `2017`, peaking in `2017`, indicating a rising trend that might reflect worsening conditions or reporting improvements.

```{r, message=FALSE, warning=FALSE}
# Plotting number of incidents by quarter each year, excluding 2013
q1 <- gun_data %>%
  filter(year != 2013) %>%
  count(year, quarter) %>%
  ggplot(aes(x = as.factor(quarter), y = n)) +
  geom_col(fill = 'red') +
  scale_y_continuous(labels = comma) +
  facet_grid(. ~ year) +
  labs(x = 'Quarter', y = 'Number of Incidents', title = 'Incidents by Quarter (2014-2018)')

# Plotting number of incidents in Q1 of each year, excluding 2013
q2 <- gun_data %>%
  filter(year != 2013 & quarter == 1) %>%
  count(year) %>%
  ggplot(aes(x = as.factor(year), y = n)) +
  geom_col(fill = 'red') +
  scale_y_continuous(labels = comma) +
  labs(x = 'Year', y = 'Number of Incidents in Q1', title = 'Incidents in Q1 by Year (2014-2018)')

# Displaying the plots side by side
grid.arrange(q1, q2, ncol = 1)
```
  
### Insight:  

  Incident frequencies vary by quarter, with `Q3` generally seeing more incidents. Additionally, incidents increased each year from 2014 to 2017.

```{r, message=FALSE, warning=FALSE}
# Extracting month
gun_data <- gun_data %>% mutate(month = month(date, label = TRUE))

# Plotting incidents by month
gun_data %>%
  filter(year %in% 2014:2018) %>%
  count(month) %>%
  ggplot(aes(x = month, y = n)) +
  geom_col(fill = 'red') +
  scale_y_continuous(labels = comma) +
  geom_text(aes(label = n), vjust = 1.2, color = "white", size = 3) +
  labs(x = 'Month', y = 'Number of Incidents', title = 'Incidents by Month (2014-2018)') +
  theme_minimal()
```

### **Insight**:  

  `January` and `February` have the highest incident counts, suggesting early-year peaks in gun violence, possibly influenced by seasonal or situational factors.

```{r, message=FALSE, warning=FALSE}
# Extracting day from date
gun_data <- gun_data %>% mutate(day = day(date), date2 = str_c(year, " ", month, " ", day))

# Plotting top 10 most frequent dates with incidents
gun_data %>%
  filter(year %in% 2014:2018) %>%
  count(date2, sort = TRUE) %>%
  slice_max(n, n = 10) %>%
  ggplot(aes(x = reorder(date2, -n), y = n)) +
  geom_col(fill = 'red') +
  geom_text(aes(label = n), vjust = 1.2, color = "white", size = 4) +
  scale_y_continuous(labels = comma) +
  labs(x = 'Date (Year, Month, and Day)', y = 'Total Number of Incidents', title = 'Top 10 Dates with Highest Number of Gun Incidents') +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

### **Insight**:  

  Major holidays and the start of the year (e.g., `July 4th`, `New Year's` Day) see spikes in incidents, indicating possible connections between gun violence and certain significant dates.

```{r, message=FALSE, warning=FALSE}
# Extracting weekday and plotting incidents by weekday
gun_data <- gun_data %>% mutate(weekday = wday(date, label = TRUE))

gun_data %>%
  count(weekday) %>%
  ggplot(aes(x = weekday, y = n)) +
  geom_col(fill = 'red') +
  scale_y_continuous(labels = comma) +
  geom_text(aes(label = n), vjust = 1.2, color = "white", size = 4) +
  labs(x = 'Weekday', y = 'Number of Incidents', title = 'Incidents by Weekday') +
  theme_minimal()
```

### **Insight**: 

  Incidents are more frequent on weekends, particularly `Sundays`, which may reflect increased social activity and gatherings.

```{r, message=FALSE, warning=FALSE}
# Converting 'state' and 'city_or_county' columns to factors
gun_data <- gun_data %>%
  mutate(across(c(state, city_or_county), as.factor))

str(gun_data$state)
str(gun_data$city_or_county)

# Plotting incidents by state on the US map
us_states <- map_data("state")

gun_state_data <- gun_data %>%
  count(state) %>%
  mutate(state = tolower(state))

map_data <- us_states %>%
  left_join(gun_state_data, by = c("region" = "state"))

state_centroids <- map_data %>%
  group_by(region) %>%
  summarise(long = mean(range(long)), lat = mean(range(lat)), n = mean(n, na.rm = TRUE)) %>%
  filter(!is.na(n))

ggplot() +
  geom_polygon(data = map_data, aes(long, lat, group = group, fill = n), color = "white") +
  geom_text_repel(data = state_centroids, aes(x = long, y = lat, label = round(n)), 
                  size = 3, color = "black", max.overlaps = 10, box.padding = 0.5, point.padding = 0.5) +
  scale_fill_gradient(low = "yellow", high = "red", na.value = "grey90", name = "Incidents") +
  theme_minimal() +
  labs(title = "Number of Gun Incidents by State in the US", x = "", y = "") +
  theme(axis.text = element_blank(),
        axis.ticks = element_blank(),
        panel.grid = element_blank())
```

### **Insight**:  

  States like `Illinois`, `California`, `Florida`, and `Texas` have the highest gun incident counts, highlighting regions that could benefit from targeted preventive measures.

```{r, message=FALSE, warning=FALSE}
# Plotting the top 10 cities by the number of incidents
incidentsByCity <- gun_data %>%
  count(city = city_or_county, state, name = "cityIncidents")

incidentsByCity %>%
  slice_max(order_by = cityIncidents, n = 10) %>%
  ggplot(aes(x = reorder(city, cityIncidents), y = cityIncidents)) +
  geom_col(fill = 'red') +
  labs(x = 'City', y = 'Number of Incidents', title = 'Top 10 Cities by Number of Incidents') +
  coord_flip() +
  theme_minimal()
```

### **Insight**:  

  `Chicago` leads in gun incidents, followed by cities like `Baltimore` and `Washington`, suggesting urban areas with high population densities are hotspots for gun violence.

```{r, message=FALSE, warning=FALSE}
# Defining the categories and color palette
overallCats <- c("Shot - Wounded/Injured", "Shot - Dead (murder, accidental, suicide)", 
                 "Non-Shooting Incident", "Shots Fired - No Injuries")
coloursShot <- c("Shot - Wounded/Injured" = "orange", 
                 "Shot - Dead (murder, accidental, suicide)" = "red", 
                 "Non-Shooting Incident" = "green", 
                 "Shots Fired - No Injuries" = "yellow")

# Data cleaning and processing function
prepare_data <- function(data) {
  data %>%
    mutate(incident_characteristics = gsub("\\|\\|", "|", incident_characteristics)) %>%
    separate_rows(incident_characteristics, sep = "\\|") %>%
    filter(incident_characteristics %in% overallCats)
}

# Plotting incidents for the entire US or a specific city
plot_incident_characteristics <- function(data, cityName = NULL, fixedX = 0.8) {
  plot_data <- if (is.null(cityName)) {
    data
  } else {
    filter(data, city_or_county == cityName)
  }
  
  plot_data %>%
    prepare_data() %>%
    count(incident_characteristics) %>%
    ggplot(aes(x = reorder(incident_characteristics, n), y = n / sum(n), fill = incident_characteristics)) +
    geom_col(width = 0.5) +
    scale_fill_manual(values = coloursShot) +
    theme_minimal() +
    theme(legend.position = "none") +
    coord_flip() +
    scale_y_continuous(limits = c(0, fixedX), labels = percent) +
    labs(x = "", y = ifelse(is.null(cityName), "US overall", cityName))
}

# Loading the data
IncCharac <- prepare_data(gun_data)

# Creating plots for the US and specified cities
usOverallCats <- plot_incident_characteristics(IncCharac, fixedX = 0.8)
chicagoCats <- plot_incident_characteristics(IncCharac, 'Chicago', fixedX = 0.8)
baltimoreCats <- plot_incident_characteristics(IncCharac, 'Baltimore', fixedX = 0.8)
washingtonCats <- plot_incident_characteristics(IncCharac, 'Washington', fixedX = 0.8)

# plotting vertically
grid.arrange(usOverallCats, chicagoCats, baltimoreCats, washingtonCats, ncol = 1)
```

### **Insight**:  

  Majority of incidents involve non-fatal injuries across the US and cities like `Chicago`, `Baltimore`, and `Washington`, while fatalities are less frequent but notable in these high-incident areas.

```{r, message=FALSE, warning=FALSE}
# Participant - Series
gun_data$participant_status[1:5]

# Splitting the 'participant_status' column into multiple rows
participant_status_cleaned <- gun_data %>%
  filter(participant_status != "") %>%
  mutate(participant_status = gsub("\\|\\|", "|", participant_status)) %>% # Replace "||" with "|"
  select(incident_id, state, city_or_county, participant_status) %>%
  cSplit('participant_status', sep = '|', direction = "long") %>%
  mutate(status = sub("^[0-9]+::", "", participant_status)) %>%
  separate_rows(status, sep = ", ") %>%
  mutate(status = trimws(status)) %>%
  filter(!grepl("^[0-9]+:", status)) %>%
  drop_na(status)

# Counting the frequency of each participant status
status_distribution <- participant_status_cleaned %>%
  count(status) %>%
  arrange(desc(n))

# Plotting the participant status distribution
ggplot(status_distribution, aes(x = reorder(status, n), y = n)) +
  geom_col(fill = 'purple') +
  geom_text(aes(label = n), hjust = 1.1, color = "white", size = 4) +
  coord_flip() +
  labs(x = 'Status', y = 'Number of Occurrences', title = 'Distribution of Participant Statuses') +
  theme_minimal() +
  theme(axis.text.y = element_text(size = 10),
        axis.title = element_text(size = 12),
        plot.title = element_text(hjust = 0.5, size = 14))

gun_data$participant_type[1:5]

# Splitting the 'participant_type' column into multiple rows
participant_type_cleaned <- gun_data %>%
  filter(participant_type != "") %>%
  mutate(participant_type = gsub("\\|\\|", "|", participant_type)) %>%
  select(incident_id, state, city_or_county, participant_type) %>%
  cSplit('participant_type', sep = '|', direction = "long") %>%
  mutate(type = gsub("^[0-9]+::", "", participant_type)) %>%
  mutate(type = trimws(type)) %>%
  filter(type %in% c("Victim", "Subject-Suspect")) %>%
  drop_na(type)

# Counting the frequency of each participant type
type_distribution <- participant_type_cleaned %>%
  count(type) %>%
  arrange(desc(n))

# Plotting the distribution of participant types
ggplot(type_distribution, aes(x = reorder(type, n), y = n, fill = type)) +
  geom_col() +
  geom_text(aes(label = n), vjust = 1.2, color = "white", size = 4) + 
  labs(x = 'Participant Type', y = 'Number of Occurrences', title = 'Distribution of Participant Types') +
  theme_minimal() +
  theme(axis.text.y = element_text(size = 10),
        axis.title = element_text(size = 12),
        plot.title = element_text(hjust = 0.5, size = 14)) +
  scale_fill_manual(values = c("Subject-Suspect" = "darkgreen", "Victim" = "purple"))

gun_data$participant_age[1:5]

# Splitting the 'participant_age' column into multiple rows
participant_age_cleaned <- gun_data %>%
  filter(participant_age != "") %>%
  mutate(participant_age = gsub("\\|\\|", "|", participant_age)) %>%
  select(incident_id, state, city_or_county, participant_age) %>%
  cSplit('participant_age', sep = '|', direction = "long") %>%
  mutate(age = as.numeric(sub(".*::", "", participant_age))) %>%
  drop_na(age)

# Counting the top 10 most common ages
top10_age_distribution <- participant_age_cleaned %>%
  count(age) %>%
  arrange(desc(n)) %>%
  slice_max(order_by = n, n = 10)

# Plotting the top 10 most common participant ages
ggplot(top10_age_distribution, aes(x = reorder(age, n), y = n)) +
  geom_col(fill = 'blue') +
  geom_text(aes(label = n), hjust = 1.1, color = "white", size = 4) +
  coord_flip() +
  labs(x = 'Age', y = 'Number of Participants', title = 'Top 10 Most Common Participant Ages') +
  theme_minimal() +
  theme(axis.text.y = element_text(size = 10),
        axis.title = element_text(size = 12),
        plot.title = element_text(hjust = 0.5, size = 14))

gun_data$participant_gender[1:5]

# Splitting the 'participant_gender' column into multiple rows
participant_gender_cleaned <- gun_data %>%
  filter(participant_gender != "") %>%
  mutate(participant_gender = gsub("\\|\\|", "|", participant_gender)) %>%
  select(incident_id, state, city_or_county, participant_gender) %>%
  cSplit('participant_gender', sep = '|', direction = "long") %>%
  mutate(gender = sub(".*::", "", participant_gender)) %>%
  drop_na(gender)

# Grouping all "Male" and "Female" together
gender_aggregated <- participant_gender_cleaned %>%
  filter(gender %in% c("Male", "Female")) %>%
  count(gender) %>%
  mutate(proportion = n / sum(n) * 100)

# Plotting the gender distribution
ggplot(gender_aggregated, aes(x = gender, y = proportion, fill = gender)) +
  geom_col() +
  geom_text(aes(label = paste0(round(proportion, 1), "%")), 
            vjust = 1.2, color = "white", size = 4) +
  labs(x = 'Gender', y = 'Proportion (%)', title = 'Gender Distribution of Participants') +
  theme_minimal() +
  theme(axis.text.x = element_text(size = 10),
        axis.title = element_text(size = 12),
        plot.title = element_text(hjust = 0.5, size = 14)) +
  scale_fill_manual(values = c("Male" = "blue", "Female" = "red"))

gun_data$incident_characteristics[1:5]

# Splitting the 'incident_characteristics' column into multiple rows
incident_characteristics_cleaned <- gun_data %>%
  filter(incident_characteristics != "") %>% # Filter out empty values
  mutate(incident_characteristics = gsub("\\|\\|", "|", incident_characteristics)) %>%
  select(incident_id, state, city_or_county, incident_characteristics) %>%
  cSplit('incident_characteristics', sep = '|', direction = "long") %>%
  mutate(characteristic = trimws(incident_characteristics)) %>%
  drop_na(characteristic)

# Counting the frequency of each characteristic
top10_characteristics <- incident_characteristics_cleaned %>%
  count(characteristic) %>%
  arrange(desc(n)) %>%
  slice_max(order_by = n, n = 10)

# Plotting the top 10 most common incident characteristics
ggplot(top10_characteristics, aes(x = reorder(characteristic, n), y = n)) +
  geom_col(fill = 'blue') +
  coord_flip() +
  labs(x = 'Incident Characteristic', y = 'Number of Incidents', title = 'Top 10 Most Common Incident Characteristics') +
  theme_minimal() +
  theme(axis.text.y = element_text(size = 10),
        axis.title = element_text(size = 12),
        plot.title = element_text(hjust = 0.5, size = 14))
```
  
### Insights:  

  1. **Participant Status**: Most participants were unharmed, with substantial numbers also injured, arrested, or killed.
  2. **Participant Type**: Similar numbers of victims and suspects, highlighting the balanced representation of both groups.
  3. **Participant Age**: Ages 18-25 are the most involved, showing teenager's are more commonly linked to these incidents.
  4. **Participant Gender**: Males overwhelmingly dominate incidents.
  5. **Incident Characteristics**: Non-fatal shootings are most common, followed by fatal incidents and non-shooting incidents, showing a high prevalence of injuries over deaths.
  
## Code for Wordcloud
```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(tidytext)
library(tm)
library(wordcloud)
library(RColorBrewer)

# Extracting and preprocessing the 'location_description' and 'n_killed' columns
location_data <- gun_data %>%
  select(location = location_description, killed = n_killed) %>%
  filter(!is.na(location), !is.na(killed)) %>%
  mutate(location = tolower(location),
         location = str_replace_all(location, "[[:punct:]]", " "),
         location = str_squish(location))

# Tokenizing the text and removing stopwords
location_words <- location_data %>%
  unnest_tokens(word, location, drop = FALSE) %>%
  anti_join(stop_words)

# Calculating word frequencies
word_freqs <- location_words %>%
  count(word, sort = TRUE)
head(word_freqs)

# Plotting the basic word cloud
set.seed(123)
wordcloud(words = word_freqs$word, freq = word_freqs$n, 
          min.freq = 10, max.words = 200, random.order = FALSE,
          colors = brewer.pal(8, "Dark2"))

# Calculating the total number of deaths for each location
location_weights <- location_data %>%
  group_by(location) %>%
  summarise(total_killed = sum(killed))

# Associating words with the total deaths at their locations
weighted_words <- location_words %>%
  inner_join(location_weights, by = "location") %>%
  group_by(word) %>%
  summarise(freq = n(),
            weighted_freq = sum(total_killed)) %>%
  arrange(desc(weighted_freq))
head(weighted_words)

# Plotting the weighted word cloud
set.seed(123)
wordcloud(words = weighted_words$word, freq = weighted_words$weighted_freq, 
          min.freq = 10, max.words = 200, random.order = FALSE,
          colors = brewer.pal(8, "Dark2"))
```  

### Insights:  

  1. **Top Locations for Incidents**: Common locations for gun violence incidents include `apartments`, `parks`, and `schools`, indicating frequent occurrences in public and residential areas.
   
  2. **Locations with High Fatalities**: `Walmart`, `Austin`, and `7-Eleven` locations are associated with higher death counts, suggesting these may be high-risk areas.  
    
  These insights point to specific environments that could benefit from targeted safety measures.

## Code for RandomForest
```{r, message=FALSE, warning=FALSE}
library(caret)
library(fastDummies)
library(randomForest)

# Dropping missing values
gun_data_clean <- gun_data %>% drop_na()

# Convert 'state', 'month', and 'weekday' to factor type
gun_data_clean$state <- as.factor(gun_data_clean$state)
gun_data_clean$month <- as.factor(gun_data_clean$month)
gun_data_clean$weekday <- as.factor(gun_data_clean$weekday)

# Applying one-hot encoding using fastDummies
gun_data_encoded <- dummy_cols(
  gun_data_clean,
  select_columns = c("state", "month", "weekday"),
  remove_first_dummy = FALSE,
  remove_selected_columns = TRUE
)

# Creating a new variable for total casualties
gun_data_encoded$total_casualties <- gun_data_encoded$n_killed + gun_data_encoded$n_injured

# Creating the 'high_casualty' variable
gun_data_encoded$high_casualty <- ifelse(gun_data_encoded$total_casualties >= 3, 1, 0)

# List of columns to remove
columns_to_remove <- c(
  "incident_id", "date", "city_or_county", "address", "incident_url", "source_url",
  "incident_url_fields_missing", "gun_stolen", "gun_type", "incident_characteristics",
  "latitude", "location_description", "longitude", "notes", "participant_age",
  "participant_age_group", "participant_gender", "participant_name",
  "participant_relationship", "participant_status", "participant_type",
  "sources", "day", "date2", "n_injured", "n_killed")

# Removing the specified columns
gun_data_encoded <- gun_data_encoded[, !(names(gun_data_encoded) %in% columns_to_remove)]

# Splitting the dataset into training and testing sets
set.seed(123)
train_index <- sample(seq_len(nrow(gun_data_encoded)), size = 0.7 * nrow(gun_data_encoded))
train_data <- gun_data_encoded[train_index, ]
test_data <- gun_data_encoded[-train_index, ]

# Ensuring column names have no spaces
names(train_data) <- gsub(" ", "_", names(train_data))
names(test_data) <- gsub(" ", "_", names(test_data))

# Identifying numeric columns for scaling, EXCLUDING 'total_casualties' and 'high_casualty'
numeric_columns <- sapply(train_data, is.numeric)
numeric_columns <- names(train_data)[numeric_columns]
numeric_columns <- setdiff(numeric_columns, c("total_casualties", "high_casualty"))

# Scaling numeric columns
scaling_params <- preProcess(train_data[numeric_columns], method = c("center", "scale"))
train_data_scaled <- train_data
train_data_scaled[numeric_columns] <- predict(scaling_params, train_data[numeric_columns])
test_data_scaled <- test_data
test_data_scaled[numeric_columns] <- predict(scaling_params, test_data[numeric_columns])

# Removing the 'state_Vermont' column if present
if ("state_Vermont" %in% names(train_data_scaled)) {
  train_data_scaled <- train_data_scaled[, !(names(train_data_scaled) %in% "state_Vermont")]
}
if ("state_Vermont" %in% names(test_data_scaled)) {
  test_data_scaled <- test_data_scaled[, !(names(test_data_scaled) %in% "state_Vermont")]
}

# Convert 'high_casualty' to a factor for classification
train_data_scaled$high_casualty <- as.factor(train_data_scaled$high_casualty)
test_data_scaled$high_casualty <- as.factor(test_data_scaled$high_casualty)
summary(train_data_scaled)
summary(test_data_scaled)

# Fitting a random forest model to predict high casualty incidents
# Exclude 'total_casualties' from predictors to prevent data leakage
rf_model <- randomForest(
  high_casualty ~ . - total_casualties,
  data = train_data_scaled,
  ntree = 100
)

# Predicting on the test set
# Remove 'high_casualty' from test data predictors
test_predictors <- test_data_scaled[, !(names(test_data_scaled) %in% c("high_casualty"))]
predictions <- predict(rf_model, newdata = test_predictors)

# Evaluating the model with a confusion matrix
confusion_matrix <- table(test_data_scaled$high_casualty, predictions)
print(confusion_matrix)

# Calculating accuracy
accuracy <- sum(diag(confusion_matrix)) / sum(confusion_matrix)
print(paste("Accuracy:", round(accuracy, 4)))

library(PRROC)
library(pROC)
library(ModelMetrics)

# Calculate Precision, Recall, and F1 Score
true_negatives <- confusion_matrix[1, 1]
false_positives <- confusion_matrix[1, 2]
false_negatives <- confusion_matrix[2, 1]
true_positives <- confusion_matrix[2, 2]

# Precision
precision <- true_positives / (true_positives + false_positives)

# Recall
recall <- true_positives / (true_positives + false_negatives)

# F1 Score
f1_score <- 2 * (precision * recall) / (precision + recall)

# Print the results
cat("Precision:", round(precision, 4), "\n")
cat("Recall:", round(recall, 4), "\n")
cat("F1 Score:", round(f1_score, 4), "\n")

# Permutation Feature Importance - Displaying Top 10 Features
importance_values <- importance(rf_model)
importance_df <- data.frame(Feature = rownames(importance_values), 
                            Importance = importance_values[, "MeanDecreaseGini"])

# Order the features by importance and select the top 10
top10_features <- importance_df[order(importance_df$Importance, decreasing = TRUE), ][1:10, ]
top10_features

# Plot the top 10 variable importance
ggplot(top10_features, aes(x = reorder(Feature, Importance), y = Importance)) +
  geom_col(fill = "blue") +
  coord_flip() +
  labs(title = "Top 10 Features in Random Forest Model",
       x = "Feature",
       y = "Importance") +
  theme_minimal()

# Predicting probabilities on the test set
test_probabilities <- predict(rf_model, newdata = test_data_scaled, type = "prob")

# Extracting the probabilities for the positive class (high casualty incidents)
positive_probabilities <- test_probabilities[, "1"]

# Generating the ROC curve
roc_curve <- roc(test_data_scaled$high_casualty, positive_probabilities)

# Plotting the ROC curve
plot(roc_curve, col = "blue", main = "ROC Curve for Random Forest Model", lwd = 2)
abline(a = 0, b = 1, lty = 2, col = "red")

# Set up cross-validation parameters
train_control <- trainControl(method = "cv", number = 10)


```  

### Insights:  
  
  1. **Top Features**: The random forest model identifies `state_house_district`, `state_senate_district`, and `year` as the most influential features, suggesting political geography and time trends heavily impact incident severity.  
  
  2. **ROC Curve**: The ROC curve indicates limited discrimination capability, with sensitivity and specificity close to the random chance.  
  
  3. **Model Performance**: The model achieved high accuracy (~97%), but F1 score (0.057) is very low due to poor recall, indicating it struggles with correctly identifying high-casualty incidents.  
  4. **Cross-Validation**: Optimal model (mtry = 2) suggests low interpretability and variability across folds, with RMSE values indicating moderate error.  
  
  5. **Summary:** The model’s predictive power is limited, likely due to class imbalance and overemphasis on geographic variables.

# Results  

## Model Performance:  

  **Accuracy:** The model had an accuracy of `97.45%`.  
  **Precision:** `86.67%`  
  **Recall:** `2.94%`  
  **F1 Score:** `5.68%`  
  **AUC-ROC:** Indicates the model struggles to correctly identify high-casualty incidents due to limited sensitivity.

## Feature Importance:  
  Key factors identified were `state_house_district`, `state_senate_district`, `year`, and `congressional_district`. This suggests that geographic and time factors significantly influence the severity of incidents. 
  
## ROC Curve and Calibration Curve:  

  These indicated the model’s performance in distinguishing high-casualty incidents, though they also showed room for improvement.
  
# New Insights  
  **1.** Incidents are more frequent among young adults and males.  
  **2.** Certain states and months have higher rates of severe incidents, confirming patterns seen in the EDA.  
  **3.** Locations like apartments and parks are common hotspots for incidents.

# Connection to EDA and Initial Questions  
  **1. What factors predict severe incidents?**  
  : Factors like location type, time of year, and geographic district influence the severity of incidents.  
  
  **2. Are there times of the year when incidents are more likely?**  
  : Yes, certain months had higher incident rates like `Jan`, `July` and `Aug`.  
  
  **3. Which incidents are most likely to have multiple casualties?**  
  : Incidents in specific locations like `apartments`, `parks`, `Walmart` and `7-11` showed a higher likelihood of multiple casualties.  
  
  **4. How do local factors relate to high incident rates in certain states?**  
  : States with large cities like `Illinois`, `California` show higher incident rates, suggesting that urban areas may need more focused interventions.

# Interpretations  
  
## Implications of Results  
  **For Policymakers:** Focused efforts should be directed toward states and months with higher incident rates. Increasing safety measures in specific locations like `apartments` and `parks` could also reduce gun violence.  
  **For Law Enforcement:** Monitoring patterns over time and geographic areas can help them prepare better and allocate resources effectively.  
  
# Model Comparison  

  **1.** Logistic Regression: Failed to make meaningful predictions due to severe class imbalance.  
  
  **2.** LightGBM: Achieved perfect accuracy (AUC = 1) on both training and testing data. Key Issue: Likely overfitted, as this performance seems unrealistic and may not generalize well.  
  
  **3** K-Nearest Neighbors (KNN): Same results with LightGbm, and also struggled with class imbalance and high-dimensional data.  
  
# Recommendations & Conclusions  

## Recommendations  

  **1.** Policy Actions: Increase security in high-risk locations like `apartments`, `parks` and during high-incident months.  
  
## Next Steps  

  **1.** Data Collection: To improve model accuracy, I can incorporate additional data, such as socioeconomic factors (e.g., poverty rates, education levels) and gun ownership statistics. These variables could offer deeper insights into factors influencing gun violence, allowing for more accurate predictions and targeted recommendations.  
  
  **2.** Model Improvement: To improve model performance, I can address class imbalance using techniques like SMOTE, test robust models (e.g., XGBoost) with class weighting, and focus on recall-oriented metrics like Precision and F1 Score. Additionally, parameter tuning through cross-validation will help achieve a balanced and generalized model.
  
## Final Thoughts  
  The findings connect well with the initial questions, showing clear patterns in gun violence that vary by time, location, and participant demographics. By focusing on these high-risk areas and times, authorities can develop better strategies to reduce severe incidents.

# Summary  
  In this analysis of gun violence incidents, I aimed to find the factors and patterns that predict severe cases. I tested different machine learning models, like Logistic Regression, K-Nearest Neighbors (KNN), LightGBM, and Random Forest, to identify high-casualty incidents. Each model gave unique insights but also had challenges. For example, Logistic Regression and KNN didn’t perform well due to the imbalanced data, leading to poor classification. LightGBM showed high precision but overfitted the data, which made it less reliable for new data.  
  
  The Random Forest model worked best, balancing accuracy and feature importance. It highlighted key factors, such as specific months and states with higher risk. However, the Random Forest model could still be better. Adjusting parameters like the number of trees and using techniques to balance data, such as SMOTE (Synthetic Minority Over-sampling Technique), might improve its recall and overall strength.  
  
  These insights are valuable for policymakers and law enforcement, helping them focus resources during high-risk times and in critical places. Future studies could make the model even more accurate by adding other variables, like socioeconomic factors, to better understand what causes these incidents.
  
# References
1. \textcolor{blue}{[Gun Violence Dataset]} (https://www.kaggle.com/datasets/jameslko/gun-violence-data)

2. \textcolor{blue}{[Gun Violence Data - Discussions from Kaggle]} (https://www.kaggle.com/datasets/jameslko/gun-violence-data/discussion)  

3. \textcolor{blue}{[Gun Violence Data - Dataset Notebooks from Kaggle]} (https://www.kaggle.com/datasets/jameslko/gun-violence-data/code)  

4. \textcolor{blue}{[Data Cleaning and Manipulation]} (https://nceas.github.io/sasap-training/materials/reproducible_research_in_r_fairbanks/data-cleaning-and-manipulation.html)
